---
title: "Predicting Diabetes"
authors:
- Demi van den Biggelaar (9660089)
- "August Gesthuizen (5292565)"
- "Friso Harff (7526946)"
- "Leander van der Waal (7180063)"
date: "2026-01-15"
format:
  html:
    embed-resources: true
---

## Inhoudsopgave - Leander

When you click the **Render** button a document will be generated that includes:

-   Content authored with markdown
-   Output from executable code

## Inlezen data + Data uitleggen + research questions - Leander

When you click the **Render** button a presentation will be generated that includes both content and the output of embedded code. You can embed code like this:

```{r }
1 + 1
```

## Beste model kiezen en uitleggen (leg uit wat het beste model is, wat het doet en waarom het het beste is) - August

When you click the **Render** button a presentation will be generated that includes both content and the output of embedded code. You can embed code like this:

```{r }
1 + 1
```

## assumptions, welke assumptions, waarom ze voldoen - Friso

## Binary dependent variable

-   Diabetes (coded 1)
-   Not diabetes (coded 0)

```{r }
# Mutating Outcome into factor (necessary for next part)
dataRDS$Outcome <- as.factor(dataRDS$Outcome)

dataRDS %>% 
  ggplot(aes(x = Outcome, fill = Outcome)) +
  geom_bar() +
  labs(x = "Outcome",
       y = "Count",
       title = "Distribution of the Outcome variable") +
  theme_bw()
# You can see that there are indeed only two outcomes for `Outcome`, so our outcome follows a binomial distribution.
```

## Sufficiently large sample size

-   10 cases per candidate predictor : $N=\frac{10k}{p}$

```{r}
#First we need to get the proportion of people that have diabetes in our sample, which is 0.33

dataRDS %>% 
  count(Outcome) %>% 
  mutate(prop = n / sum(n))
```

```{r}
#Now we can plug this into our formula to get the minimum number of positive cases. 
ssize_cal <- function(k, p){
  round((10*k)/p)
}

ssize_cal(2, 0.33)
```

## Full-rank predictor matrix

-   More observations than predictors

-   No multicollinearity among linear predictors

```{r}
VIF(fit1)
```

## assumptions, delete outliers, laat zien dat ie niet van belang is op case by case basis - Friso

When you click the **Render** button a presentation will be generated that includes both content and the output of embedded code. You can embed code like this:

```{r }
1 + 1
```

## Exploring the model section, ga in op het model met onderzoeksmethode uit slides INCLUDING CONFUSION MATRIX - Demi

When you click the **Render** button a presentation will be generated that includes both content and the output of embedded code. You can embed code like this:

```{r }
1 + 1
```

## Concludeer en beantwoord - Demi

When you click the **Render** button a presentation will be generated that includes both content and the output of embedded code. You can embed code like this:

```{r }
1 + 1
```
