---
title: "Predicting Diabetes"
authors:
- Demi van den Biggelaar (9660089)
- "August Gesthuizen (5292565)"
- "Friso Harff (7526946)"
- "Leander van der Waal (7180063)"
date: "2026-01-15"
format:
  html:
    embed-resources: true
---

## Intro

authors:
- Demi van den Biggelaar (9660089)
- August Gesthuizen (5292565)
- Friso Harff (7526946)
- Leander van der Waal (7180063)

## Inhoudsopgave

When you click the **Render** button a document will be generated that includes:

-   Content authored with markdown
-   Output from executable code

## Inlezen data + research questions

When you click the **Render** button a presentation will be generated that includes both content and the output of embedded code. You can embed code like this:

```{r }
1 + 1
```

## Beste model kiezen en uitleggen (leg uit wat het beste model is, wat het doet en waarom het het beste is)

When you click the **Render** button a presentation will be generated that includes both content and the output of embedded code. You can embed code like this:

```{r }
1 + 1
```

## assumptions, welke assumptions, waarom ze voldoen 

When you click the **Render** button a presentation will be generated that includes both content and the output of embedded code. You can embed code like this:

```{r }
1 + 1
```

## assumptions, delete outliers, laat zien dat ie niet van belang is op case by case basis
When you click the **Render** button a presentation will be generated that includes both content and the output of embedded code. You can embed code like this:

```{r }
1 + 1
```

## Exploring the model section, ga in op het model met onderzoeksmethode uit slides INCLUDING CONFUSION MATRIX
When you click the **Render** button a presentation will be generated that includes both content and the output of embedded code. You can embed code like this:

```{r }
1 + 1
```

## Concludeer en beantwoord
When you click the **Render** button a presentation will be generated that includes both content and the output of embedded code. You can embed code like this:

```{r }
1 + 1
```
