---
title: "Predicting Diabetes"
authors:
- Demi van den Biggelaar (9660089)
- "August Gesthuizen (5292565)"
- "Friso Harff (7526946)"
- "Leander van der Waal (7180063)"
date: "2026-01-15"
format:
  html:
    embed-resources: true
---

# **Prediction of glucose using linear regression**

Demi van den Biggelaar (9660089)\
August Gesthuizen (5292565)\
Friso Harff (7526946)\
Leander van der Waal (7180063)\

## **Introduction**

The different parameters in the dataset are:

-   Pregnancies: Number of times pregnant

-   Glucose: Plasma glucose concentration after 2 hours in an oral glucose tolerance test

-   BloodPressure: Diastolic blood pressure (mm Hg)

-   SkinThickness: Triceps skin fold thickness (mm)

-   Insulin: 2-Hour serum insulin (mu U/ml)

-   BMI: Body mass index (weight in kg/(height in m)\^2)

-   DiabetesPedigreeFunction: Diabetes pedigree function

-   Age: Age (years)

-   Outcome: Class variable (0 or 1, where 1 means positive for diabetes)

## **Preprocessing & Loading data**

To preprocess the data we categorized the BMI into four categories, being underweight, healthy, overweight and obese.

------------------------------------------------------------------------

```{r}
# Initialization
setwd(".")
library(ggplot2)
library(car)
```

```{r}
#| warning: false

# Load .CSV and save as .RDS

datacsv <- read.csv("./data/raw/diabetes.csv", header = TRUE)
saveRDS(datacsv, "./data/raw/dataset.rds")

# Preprocessing
dataRDS <- readRDS("./data/raw/dataset.rds")
dataRDS <- na.omit(dataRDS) # removes NA from important columns
saveRDS(dataRDS, "./data/processed/dataset.rds")
dataset <- readRDS("./data/processed/dataset.rds")

# Recoding BMI to classes
dataRDS$BMI[dataRDS$BMI <= 18.5 ] <- 1
dataRDS$BMI[dataRDS$BMI > 18.5 & dataRDS$BMI <= 25  ] <- 2
dataRDS$BMI[dataRDS$BMI > 25 & dataRDS$BMI <= 30 ] <- 3
dataRDS$BMI[dataRDS$BMI > 30] <- 4

# Removing missing values
dataRDS <- subset(dataRDS,
                     Glucose != 0 &
                     Insulin != 0 &
                     BMI != 0 &
                     SkinThickness != 0)

# Adding moderation
dataRDS$moderation <- dataRDS$BMI * dataRDS$Insulin * dataRDS$SkinThickness * (dataRDS$BloodPressure) 

summary(dataset)
```

## **Exploring the data**

```{r}
#| warning: false

dataRDS
for (i in 1:ncol(dataRDS)) {
  col_name <- colnames(dataRDS)[i]
  print(hist(dataRDS[, i],
       main = paste("Histogram of ", col_name),
       xlab = paste("Values of", col_name),
       col = "lightblue",
       border = "white"))
}
pairs(dataRDS)
```
