---
title: "Main"
format: html
---

```{r}
setwd(".")
library(ggplot2)
# install.packages("pairsD3")
require("pairsD3")
```

------------------------------------------------------------------------

1.  **Preliminaries:** Introduction of your research questions, description and potential processing of your data.

------------------------------------------------------------------------

```{r}
# Load .CSV and save as .RDS

datacsv <- read.csv("./data/raw/diabetes.csv", header = TRUE)
saveRDS(datacsv, "./data/raw/dataset.rds")

# Preprocessing
dataRDS <- readRDS("./data/raw/dataset.rds")
dataRDS <- na.omit(dataRDS) # removes NA from important columns
saveRDS(dataRDS, "./data/processed/dataset.rds")
dataset <- readRDS("./data/processed/dataset.rds")

# Recoding BMI to classes
dataRDS$BMI[dataRDS$BMI <= 18.5 ] <- 1
dataRDS$BMI[dataRDS$BMI > 18.5 & dataRDS$BMI <= 25  ] <- 2
dataRDS$BMI[dataRDS$BMI > 25 & dataRDS$BMI <= 30 ] <- 3
dataRDS$BMI[dataRDS$BMI > 30] <- 4

# Adding moderation
dataRDS$moderation <- dataRDS$BMI * dataRDS$Insulin * dataRDS$SkinThickness * (dataRDS$BloodPressure) 

summary(dataset)
```

1.  **Model estimation:** Description of the model estimates, model fit, and model comparison procedure (including moderation)

```{r}
dataRDS
#shinypairs(dataRDS)
for (i in 1:ncol(dataRDS)) {
  col_name <- colnames(dataRDS)[i]
  print(hist(dataRDS[, i],
       main = paste("Histogram of ", col_name),
       xlab = paste("Column", i),
       col = "lightblue",
       border = "white"))
}
pairs(dataRDS)
```

Above you can see multiple ways that our variables are described. They are represented in two ways: using histograms and using.a panel. In the histograms you can see which values are representative of each variable while in the panel you can see the relation between of each variable with each other variable. The histograms will be briefly discussed, however remaining discussion will focus only on the relevant plots from the panel which we use for regression.

**Pregnancies:** number of times pregnant, represented as an integer, values range between 0-14

**Glucose:** plasma glucose concentration in a toleration test represented as an integer, values range from 0-200.

**BloodPressure:** Bloodpressure represented as an integer with values in the range of 0-110.

**Skinthickness:** Thickness of the skin in millimetres, represented as an integer with values ranging from 0-60.

**Insulin:** “2 hour serum insulin” with values ranging from 0-600.

**BMI:** BMI value represented as classification, namely: underweight=1, healthy=2, overweight=3 and obese=4. Most people in this study are obese.

**DiabetesPedigreeFunction:** Quantified genetic risk of diabetes. Risks range from 0-1,75. 

**Age:** Integer representing age, values range from 20-70.

**Outcome:** 0 or 1 value representing if someone has diabetes or not. Most participants don’t have diabetes.

**Moderation:** Product of the following values: BMI, Insulin, SkinThickness and BloodPressure. This value is added to support the regression accuracy.

```{r}
model <- lm(Glucose ~ Age + Insulin + BMI + Pregnancies + BloodPressure + SkinThickness + DiabetesPedigreeFunction + moderation, 
             data = dataRDS)
model2 <- lm(Glucose ~ Age + Insulin + BMI + SkinThickness + moderation + DiabetesPedigreeFunction, 
             data = dataRDS)
model3 <- lm(Glucose ~ Age + Insulin + BMI + SkinThickness + moderation, 
             data = dataRDS)

# Ask for summary statistics again
summary(model)
summary(model2)
summary(model3)

anova(model, model2, model3)


# Plot with regression line for x1
ggplot(dataRDS, aes(x = Age, y = Glucose)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) + # se = FALSE to hide confidence interval
  labs(title = "Relationship between Age and Glucose")

# Plot with regression line for x2
ggplot(dataRDS, aes(x = Insulin, y = Glucose)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Relationship between {Insulin} and Glucose")

# Plot with regression line for x1
ggplot(dataRDS, aes(x = SkinThickness, y = Glucose)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) + # se = FALSE to hide confidence interval
  labs(title = "Relationship between Age and Glucose")

# Plot with regression line for x2
ggplot(dataRDS, aes(x = BMI, y = Glucose)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Relationship between {BMI} and Glucose")

# Plot with regression line for x2
ggplot(dataRDS, aes(x = moderation, y = Glucose)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Relationship between {Moderation} and Glucose")


```

1.  **Assumptions:** Testing of model assumptions, checking for influential cases. Act upon and/or reflect on violations when needed.

```{r}
# Assumptions are: linearity, Exogenous predictors, constant finite error variance Predictor matrix full rank, independent errors, normal distributed errors.
```

1.  **Interpretation:** Substantive interpretation of the final model. Answering your research question.

```{r}

```

1.  **Layout:** Structure of the document, efficiency of output presentation, use of custom functions (when applicable). Presentation of suitable visualizations.

```{r}

```

1.  **Coding:** Use code from the labs and annotate what each piece of code does.
